Inferring Continuous and Discrete Population Genetic Structure Across Space

### 2017-08-07 writeup/conStruct.tex
merged graham's appendectomy

\textsuperscript{a}bradburd@msu.edu; 
\textsuperscript{c}gmcoop@ucdavis.edu;
\textsuperscript{b}plr@uoregon.edu\\\\\

@@

\paragraph{Relationship of this model to non-spatial structure models}

A nice feature of our approach is that the model described in Eqn \refeq{cross_cluster_covariance} 
contains a non-spatial assignment model as a special case 
(see \ref{model_app} for a more in-depth discussion). 
Setting $\alpha_0^{(k)}$ to zero for all $k$, 
we obtain a non-spatial model in which each cluster has its own allele frequency at each SNP, 
and individuals draw a proportion of their ancestry from each cluster. 
This model is strongly conceptually similar to that of STRUCTURE \citep{STRUCTURE} 
and related models \citep[e.g][]{ADMIXTURE}; 
the only difference is that our likelihood assumes that allele frequencies are normally distributed 
around their expectations, 
while the standard assignment methods assume that the error is binomial distributed (see XXXX).  
Therefore, we can evaluate the fit of the different models 
--- spatial vs. nonspatial, across different values of $K$ --- 
by comparing their performance in a common framework.

\paragraph{Choice of cluster number and cross-validation}
There are a number of reasons, both statistical and biological, 
why there is no true (or right) number of clusters for real datasets. 
We discuss these reasons further in the Discussion below. 
In practice, it is useful to explore the implications of clustering at a variety of different levels (
i.e. across multiple $K$) for biological interpretation and downstream analyses. 
However, we also recognize that in many circumstances it is 
useful to know the number of clusters with highest statistical
support in the (current) dataset.
Toward that end, we have implemented a method for 
statistically comparing \texttt{conStruct} results across different values of $K$ 
and between the spatial and non-spatial models.

@@

\section{Model rationale}
\subsection{Drift, admixture, and allele frequencies}
\paragraph{Drift} 
We imagine that the allele frequencies at each locus in each sample are the sum of 3 components: 
the ancestral allele frequency $\epsilon$ shared by all samples,
the deviation from that ancestral mean in the $k$th population, $\Delta^{(k)}$,
which is shared by all samples with 100\% ancestry in that population,
the deviation specific to the $i$th sample, $\Delta^{(i)}$,
which captures drift not shared by all samples at the population level
(i.e., subpopulation-specific drift due to, e.g., inbreeding).

If all samples drew all of their ancestry from any one of the $K$ clusters,
the  allele frequency in the $i$th sample at the $\ell$th locus would be given by:
\begin{equation}
F_{i,\ell} = \epsilon_{\ell} + \Delta^{(k)}_{\ell} + \Delta^{(i)}_{\ell}
\label{drift_terms_no_admix}
\end{equation}

\paragraph{Admixture} 
The model above describes the simple case 
in which samples draw 100\% of their ancestry from only a single cluster each. 
To accommodate admixture between clusters, 
we can model allele frequencies within samples as linear combinations of the population-specific 
drift terms across the clusters from which they draw ancestry.
To do so, we introduce a term $w^{(k)}_{i}$ that describes the admixture proportion of sample $i$ in cluster $k$,
which can be interpreted as the probability that an allele sampled in the $i$th sample came from the $k$th cluster.
The allele frequency in the $i$th sample at the $\ell$th locus can therefore be written as:
%
\begin{equation}
F_{i,\ell} = \epsilon_{\ell} + \sum\limits_{K} \left( w^{(k)}_{i} \Delta^{(k)}_{\ell} \right) + \Delta^{(i)}_{\ell}	
\label{drift_terms_admix}
\end{equation}
%
where
%
\begin{equation}
\sum\limits_{K}w^{(k)}_{i} = 1
\end{equation}

\subsection{The Normal approximation to drift}
Genetic drift is serial binomial sampling, 
but over short timescales, and for intermediate ancestral allele frequencies, 
we can reasonably model the allele frequencies at each locus across populations as multivariate normal.
This will be a good approximation if the amount of elapsed drift in each sample is relatively small,
as would be expected if time since divergence from a common ancestral population is short,
or effective population sizes are large.

Beginning with the simplest case of two clusters ($K=2$),
with all samples drawing 100\% of their ancestry from either one or the other cluster, 
we can model allele frequencies at locus $\ell$ as:
%
\begin{equation}
F_{\ell} \sim MVN\left(\mu = \epsilon_{\ell} + \iota^{(k)}\Delta_{\ell}^{(k)}, \Sigma = \Omega	\right)
\label{MVN_mean}
\end{equation}
%
where $\iota^{(k)}$ is an indicator variable that equals 1 for samples with 100\% membership in cluster $k$
and 0 otherwise.
In Eqn \eqref{MVN_mean}, all entries of the covariance $\Omega$ have value 0 except the diagonals,
which describe sample-specific drift, and have the following value:
%
\begin{equation}
\Omega_{i,i} = \text{Var}\left( \Delta^{(i)} \right)
\label{MVN_mean_diag}
\end{equation}

%

\subsection{Describing relatedness as a covariance}
However, our data can consist of many loci, 
so it may be computationally impractical to model 
the ancestral, cluster, and sample components of each sampled allele frequency.
Instead, we can describe the covariance induced between samples due to the sharing of the components.
All samples share the same ancestral mean at all loci, 
which leads to a global covariance equal to the variance of the ancestral frequencies across loci.
Likewise, samples that are in the same cluster will have a covariance equal to the variance 
of the cluster-specific deviates from the ancestral frequencies across loci.

In Eqn \ref{MVN_mean} above, we described the relatedness between samples
due to the shared ancestral allele frequency and shared population-level drift
via the mean of the multivariate normal.
We can describe this relatedness as a covariance as follows:

\begin{equation}
	\Omega_{i,j} = 
	\begin{cases}
	      \text{Var}(\epsilon) +  \text{Var}\left( \Delta^{(k)} \right) + \text{Var}\left( \Delta^{(i)} \right), & \text{if}\ i=j\ \\
      	      \text{Var}(\epsilon) +  \text{Var}\left( \Delta^{(k)} \right), & \text{if}\ \iota^{(k)}_i = \iota^{(k)}_j\ \text{and}\ i \neq j \\
	      \text{Var}(\epsilon), & \text{otherwise}
	\end{cases}
\label{MVN_covar}
\end{equation}

Generalizing this result to any number of clusters and incorporating admixture as in Eqn \ref{drift_terms_admix}, 
we can write the form of the covariance of samples that are admixed between $K$ clusters as:

\begin{equation}
\Omega_{i,j} = \text{Var}(\epsilon) + \sum\limits_K \left(	w^{(k)}_iw^{(k)}_j \text{Var}(\Delta^{(k)}) 	\right) + 
\delta_{i=j} \text{Var}(\Delta^{(i)})
\label{admixed_discrete_covariance}
\end{equation}

where $\delta_{i=j}$ is an indicator variable that equals 1 when $i$ is equal to $j$ and 0 otherwise,
as in Eqn \ref{cross_cluster_covariance}.

\subsection{Incorporating spatial differentiation}
Equation \ref{admixed_discrete_covariance} describes a model in which 
samples can be continuously admixed between a set of $K$ discrete clusters.
In this model, any pair of samples with 100\% ancestry in a cluster 
have exactly the same covariance with each other (namely $\text{Var}(\epsilon) + \text{Var}(\Delta^{(k)})$).
However, we also wish to model continuous decay of covariance with spatial separation between samples.
That is, we expect (and want our model to reflect) that samples within the same cluster 
will have higher covariance if they are sampled closer together than if they are sampled farther apart.
To describe this spatial pattern, we build in a spatial component to our covariance model.

Specifically, we write that the covariance between a pair of samples $i$ and $j$
that draw all their ancestry from a single cluster $k$ decays exponentially as a function 
of the distance between their sampling locations as follows:
\begin{equation}
G^{(k)}_{i,j} = \alpha^{(k)}_0 \times \left(\text{exp} \left(  -(\alpha^{(k)}_D D_{i,j})^{\alpha^{(k)}_2}\right) \right) + \mu^{(k)}	
\label{spatial_cov}
\end{equation}
where the $\alpha$ parameters control: 
the sill of the covariance ($\alpha_0$), 
the rate at which covariance decays with distance $D$ ($\alpha_D$),
and the shape of that decay ($\alpha_2$)
in the $k$th cluster.
The parameter $\mu^{(k)}$ describes the covariance shared by all samples in the $k$th cluster;
it and the spatial covariance function $G$ together
 describe the quantity $\text{Var}(\Delta^{(k)})$ from Eqn \eqref{admixed_discrete_covariance}.

Tying these ideas together, we can then construct a covariance that describes
\begin{enumerate}
\item continuous decay of covariance within a cluster
\item a discrete amount of covariance shared within a cluster
\item continuous admixture between clusters.
\end{enumerate}

\begin{equation}
\Omega_{i,j} = \text{Var}(\epsilon) + \sum\limits_K \left(	w^{(k)}_iw^{(k)}_j G^{(k)}_{i,j}(\theta) 	\right) +
\delta_{i=j} \text{Var}(\Delta_i)
\label{admixed_spatial_cov}
\end{equation}
%
where $G^{(k)}_{i,j}(\theta)$ denotes the dependence of the spatial covariance between samples $i$ and $j$ 
within a cluster on other quantities $\theta$: 
specifically, the parameters $\vec{\alpha}^{(k)}$ and $\mu^{(k)}$, which are specific to the $k$th cluster, 
as well as on the observed quantity $D_{i,j}$, the pairwise distance between samples $i$ and $j$.

\section{Likelihood}
If the allele frequency data are well approximated by a Gaussian, 
their sample covariance is a sufficient statistic,
so that calculating the likelihood of their sample covariance is the same as 
calculating the probability of the frequency data up to a constant. 
We can therefore model the covariance of the sample allele frequencies, $\widehat{\Omega}$, 
as a draw from a Wishart distribution with degrees of freedom equal to 
the number of loci $L$ across which the sample covariance is calculated:

\begin{align}
\widehat{\Omega} &= \frac{1}{L}FF^T \notag \\
\widehat{\Omega} &\sim \mathcal{W}\left( L\Omega, L	\right)
\label{wishart}
\end{align}

@@

\paragraph{Single cluster}
Each of these models can be run with a single cluster ($K=1$), 
in which case the cluster-specific covariance parameter $\mu^{(k})$ 
and the global covariance parameter $\gamma$ become redundant.
The single-cluster model is therefore a special case of the multi-cluster model, 
in which we set $\mu$ to zero, as described below:

>>>>

\textsuperscript{a}bradburd@msu.edu; 
\textsuperscript{b}gmcoop@ucdavis.edu;
\textsuperscript{c}plr@uoregon.edu\\\\\

@@

\paragraph{Relationship of this model to nonspatial structure models}

A nice feature of our approach is that the model described in Eqn \refeq{cross_cluster_covariance} 
contains a nonspatial assignment model as a special case 
(see \ref{model_app} for a more in-depth discussion). 
Setting $\alpha_0^{(k)}$ to zero for all $k$, 
we obtain a nonspatial model in which each cluster has its own allele frequency at each SNP, 
and individuals draw a proportion of their ancestry from each cluster. 
This model is strongly conceptually similar to that of STRUCTURE \citep{STRUCTURE} 
and related models \citep[e.g][]{ADMIXTURE}; 
the only difference is that our likelihood assumes that allele frequencies are normally distributed 
around their expectations, 
while the standard assignment methods assume that the error is binomial distributed (see XXXX).  
Therefore, we can evaluate the fit of the different models 
--- spatial vs. nonspatial, across different values of $K$ --- 
by comparing their performance in a common framework.

\paragraph{Choice of cluster number and cross-validation}
There are a number of reasons, both statistical and biological, 
why there is no true (or right) number of clusters for real datasets. 
We discuss these reasons further in the Discussion below. 
In practice, it is useful to explore the implications of clustering at a variety of different levels (
i.e. across multiple $K$) for biological interpretation and downstream analyses. 
However, we also recognize that in many circumstances it is 
useful to know the number of clusters with highest statistical
support in the (current) dataset.
Toward that end, we have implemented a method for 
statistically comparing \texttt{conStruct} results across different values of $K$ 
and between the spatial and nonspatial models.

@@

\section{Model rationale}
\subsection{Drift, admixture, and space}
\paragraph{Drift} 
We first provide a simple model of allele frequencies within a cluster. 
Imagine a sample $i$ that draws all of its ancestry from cluster $k$.
The allele frequency in sample $i$ at locus $\ell$, $F_{i,\ell}$ can be
written as the sum
\begin{equation}
F_{i,\ell} = \epsilon_{\ell} + \Delta^{(k)}_{\ell} +
\Delta^{(k,i)}_{\ell} + \Delta^{(i)}_{\ell}
\label{drift_terms_no_admix}
\end{equation}
The first term is the ancestral allele frequency $\epsilon_\ell$ shared by all samples; 
the second is the deviation from that ancestral frequency 
due to drift in the ancestral population of the $k$th cluster,
which is shared by all samples within the cluster. 
The third term is the deviation of the $i$th sample away from the $k$th cluster mean 
due to the spatial process of drift and migration within the cluster.
The final term is the deviation specific to the $i$th sample,
which captures drift not shared by all samples at the population level
(i.e., subpopulation-specific drift due to, e.g., inbreeding). 
We will assume that these four deviations are all uncorrelated with each other.

If we have two samples $i$ and $j$ drawn from cluster $k$, 
their covariance across loci will be 
\begin{equation}
\text{Var}(\epsilon) +  \text{Var}\left( \Delta^{(k)} \right) +
\text{Cov}(\Delta^{(k,i)}_{\ell},\Delta^{(k,j)}_{\ell}) + \delta_{i=j} \text{Var}\left( \Delta^{(i)} \right)\ \\
\end{equation}
where the quantity $\delta_{i=j}$ is an indicator variable that equals 1 when $i$ is equal to $j$ and 0 otherwise, 
as in Eqn. \eqref{cross_cluster_covariance}.

\paragraph{Admixture} 
The model above describes the simple case 
in which samples draw 100\% of their ancestry from only a single cluster each. 
To accommodate admixture between clusters, 
we model allele frequencies within samples as linear combinations of the population-specific 
drift terms across the clusters from which they draw ancestry.
To do so, we introduce a term $w^{(k)}_{i}$ that describes the
admixture proportion of sample $i$ in cluster $k$ ($\sum_{K}w^{(k)}_{i} = 1$),
which can be interpreted as the proportion of the genome in the $i$th
sample that came from the $k$th cluster 
(or the probability that an allele at a locus is drawn from $k$).
The allele frequency in the $i$th sample at the $\ell$th locus can therefore be written as:
\begin{equation}
F_{i,\ell} = \epsilon_{\ell} + \sum\limits_{K} w^{(k)}_{i}\left( 
  \Delta^{(k)}_{\ell} + \Delta^{(k,i)}_{\ell}\right) + \Delta^{(i)}_{\ell}	
\label{drift_terms_admix}
\end{equation}
and the covariance between $i$ and $j$ across loci is
\begin{equation}
\Omega_{i,j} = \text{Var}(\epsilon) + \sum\limits_K w^{(k)}_iw^{(k)}_j
\left(
  \text{Var}\left( \Delta^{(k)} \right) +
\text{Cov}(\Delta^{(k,i)}_{\ell},\Delta^{(k,j)}_{\ell}) 	\right) +
\delta_{i=j} \text{Var}(\Delta_i)
\label{admixed_spatial_cov}
\end{equation}

\paragraph{Space}
Under our nonspatial model, 
we assume that $\text{Cov}(\Delta^{(k,i)}_{\ell},\Delta^{(k,j)}_{\ell})=0$,
so that the only additional covariance between $i$ and $j$ 
(above that induced by a shared ancestral frequency at each locus)
is due to the drift in the ancestral population of their cluster 
(the variance of which is $\mu_k$). 
Under our spatial model we assume that some of the covariance in allele frequencies
between $i$ and $j$ decays as a function of the geographic distance
between the pair, $D_{i,j}$,
so that 
\begin{equation}
\text{Cov}(\Delta^{(k,i)}_{\ell},\Delta^{(k,j)}_{\ell}) = \alpha^{(k)}_0 \times \left(\text{exp} \left(  -(\alpha^{(k)}_D D_{i,j})^{\alpha^{(k)}_2}\right) \right)
\end{equation}
We note that this form is chosen for its flexibility, 
and not because it matches any explicit population genetic model of isolation by distance.

\paragraph{Likelihood}
If our allele frequency deviations are well approximated by a Gaussian, 
their sample covariance is a sufficient statistic,
so that calculating the likelihood of their sample covariance is the same as 
calculating the probability of the frequency data up to a constant. 
We can therefore model the covariance of the sample allele frequencies, $\widehat{\Omega}$, 
as a draw from a Wishart distribution with degrees of freedom equal to 
the number of loci $L$ across which the sample covariance is calculated:

\begin{align}
&\widehat{\Omega}_{i,j} = 
	\begin{cases}
		\frac{1}{L-1}
		\sum\limits_{\ell=1}^{L}{\left[ 	\left(f_{i,\ell} - \bar{f}_{i} \right)
		 	\left(f_{j,\ell} - \bar{f}_{j} \right)\right] \notag
			} & \text{for } i \neq j \\
		\bar{f}_i (1-\bar{f}_i) & \text{for } i = j \\
	\end{cases}\\
&\widehat{\Omega} \sim \mathcal{W}\left( L\Omega, L	\right)
\label{wishart}
\end{align}

@@

\paragraph{Single cluster}
Each of these models can be run with a single cluster ($K=1$), 
in which case the cluster-specific covariance parameter $\mu^{(k)}$ 
and the global covariance parameter $\gamma$ become redundant.
The single-cluster model is therefore a special case of the multi-cluster model, 
in which we set $\mu$ to zero, as described below:
